\ttitle

### 2015-06-06 Chapters/BodyComp.tex
Chapter Body Composition - numerous small changes - fixed all VO2 to correct format - fixed some problems with style - fixed a few spelling/grammar issues - needs a final check - see comments at end of discussion - consider separate tables for each phase of exercise

\clearpage
%----------------------------------------------------------------------------------------

\section{Introduction}
Major abdominal surgery especially for pancreatic disease is associated with significant morbidity and mortality. Patient selection is as important as identifying surgical treatable pathology in ensuring optimal outcomes. \parencite{balthazar_acute_2002}

\subsection{Role of preoperative CPET}
The role of cardiopulmonary exercise testing in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery has become well established. A number of studies have shown that poor aerobic fitness demonstrated by a low anaerobic threshold or low peak VO$_2$ or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery.\parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011} CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
Patients are sometimes denied surgery if their performance at cardiopulmonary exercise testing is felt to be poor based on currently available evidence.

@@

\subsection{Factors influencing aerobic fitness}
A low anaerobic threshold and/or low peak VO$_2$ have universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems to increased oxygen demand during exercise. This is often thought to be due to cardiorespiratory disease either overt or subclinical. Occasionally other factors such as anaemia, peripheral vascular disease and rarely mitochondrial diseases have been recognised as factors contributing to low anaerobic threshold/peak VO$_2$ or abnormalities in other parameters measured at cardiopulmonary exercise testing but this is uncommon in patients undergoing major abdominal surgery.

The most common parameters used to quantify perioperative risk in surgical patients are oxygen consumption at the anaerobic threshold (VO$_2$AT) and at peak exercise capacity (VO$_2$Peak). Conventionally these have been reported as per weight ratios in mls/kg/min. However, numerous studies on cardiorespiratory exercise physiology have reported that normalising VO$_2$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects.\parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}

\subsection{Aims}
In chapter \ref{ch_cpet_outcomes}, we reported that low anaerobic threshold in patients undergoing pancreaticoduodenectomy was associated with an increased incidence of postoperative pancreatic fistula and prolonged hospital stay. We also reported that patients with a VO$_2$AT less than 10mls/kg/min were less likely to receive postoperative adjuvant chemotherapy as a result of postoperative complications, prolonged hospital stay and likely due to lack of physiological reserve post-surgery to be fit enough to undergo chemotherapy.

However, we noted that high BMI was associated with a low VO$_2$AT independent of all other clinicopathological characteristics. Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain the very low levels of VO$_2$AT.
The aim of the present study was to explore the association between body composition, total body weight and the physiological parameters measured at cardiopulmonary exercise testing.

@@

\subsection{Patients}
Patients who underwent major abdominal surgery for malignant or benign disease involving the head of the pancreas and periampullary region at a single institution between August 2008 and October 2010 were included in this study. All data were recorded in a prospectively maintained database. Data was collected on demographics, preoperative clinicopathological characteristics including blood tests, body mass index, weight, height and the underlying surgical pathology. Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database. A detailed description of methodology of cardiopulmonary exercise testing and a description of the measured parameters is provided in CHAPTERX.

\subsection{Body composition calculation}
Preoperative computed tomography that had been performed as part of the routine assessment of these patients was used to calculate body composition. Previously published and well established methods were used were used to calculate body composition information from single CT slices.\parencite{bredella_comparison_2010,shen_total_2004}

The coronal and saggital reconstructions were used to accurately identify the L3 and L4 vertebrae.The cross-sectional images at these levels where then exported as bitmap images with C40 W350 settings [speak to a radiologist about what these numbers mean]. The scale in millimeters was included with every image. A representative image is shown in Fig. 1.The GNU Image Manipulation Program (GIMP), an advanced, free, open-source, raster graphics editor was used for analysis of all images (www.gimp.org). The use of GIMP to analyse cross-sectional imaging for body composition has been described previously although by using a different technique to what has been employed by us. \parencite{anblagan_measurement_2013}

The first step involved converted the bitmap images into JPEG images using lossy compression set at 85\% to minimise sharp transitions between grey areas of very similar colour values. This aided easier automatic selection of contiguous areas of similar grey shades.

@@

\textbf{Body compartment selection methodology}:

The sequence of steps is depicted in Fig. \ref{fig:bc_ct_gimp} on p	\pageref{fig:bc_ct_gimp}. The total cross-sectional area of the abdomen at the level of L3/L4 was calculated by first selecting all the empty space outside the image followed by inverting this selection. This is depicted in Fig. \ref{fig:bc_ct_csa}. Subcutaneous fat in the image was selected using the Fuzzy Select tool (if necessary by choosing multiple times and removing any unnecessary areas) as depicted in Fig. \ref{fig:bc_ct_sat}. The same process was repeated for visceral adipose tissue and skeletal muscle as depicted in Fig. \ref{fig:bc_ct_vat} and Fig. \ref{fig:bc_ct_sm} respectively. Every selection was visually confirmed for anatomical accuracy by using the layer selection tool to inspect the area under selection as shown in the insets in each of the images.
%CT slices images x 4
\begin{figure}[htbp]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_csa}
		\caption{Total Cross-sectional Area}
		\label{fig:bc_ct_csa}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_sat}
		\caption{Subcutaneous Adipose Tissue$^*$}
		\label{fig:bc_ct_sat}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_vat}
		\caption{Visceral Adipose Tissue$^*$}
		\label{fig:bc_ct_vat}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_sm}
		\caption{Skeletal Muscle$^*$}
		\label{fig:bc_ct_sm}
	\end{subfigure}

@@

\subsection{Cardiopulmonary exercise testing}
All patients performed cardiopulmonary exercise testing on a cycle ergometer as described in chapterx. Raw data of all breath-by-breath parameters averaged every 10 seconds was collected for analysis. The first three minutes of the recorded data were during the rest period when the patients were on the exercise bike but did not do exercise. The average of each parameter measured between the first and second minute was treated as the rest value. Anaerobic threshold was identified using previously established methods. \parencite{beaver_new_1986,sue_metabolic_1988} Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.

@@

All analyses were performed using the SPSS statistical package for Microsoft Windows (version 22 ). Comparisons between body composition and cardiopulmonary exercise testing parameters were done using the partial correlations controlling for the effect of gender (and/or age). All p-values reported are two-sided. The relationship between body composition and various preoperative clinico-pathological characteristics (in the form of categorical variables) was analysed using the Mann-Whitney U test for variables with two categories and the Kruskal-Wallis Test for variables with more than two categories. Previously established cut-offs were used for categorising continuous variables where applicable. The level of significance was set at $p<0.05$.

@@

The correlations between cardiopulmonary exercise parameters and body composition were adjusted for gender. Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to VO$_2$AT or VO$_2$Peak and therefore no adjustments were made for age. The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).

@@

Tidal volume (litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise. There was a statistically significant but weak positive correlation between Minute Ventilation (Tidal Volume x Respiratory Rate) and skeletal muscle at anaerobic threshold and peak exercise but not at rest. There was no correlation between either of these measures of pulmonary function and total adipose tissue area at any phase of exercise.

%----------------------------------------------------------------------------------------------
% Correlation between body composition and VO$_2$AT before and after correction for total body weight.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal}
		\caption{VO$_2$AT vs. Skeletal Muscle}
		\label{fig:bc_scatter_VO2_skeletal}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT}
		\caption{VO$_2$AT vs. Total Adipose Tissue}
		\label{fig:bc_scatter_VO2_TAT}
	\end{subfigure}
	\caption{Correlation between body composition and VO$_2$AT before and after correction for total body weight.}	
	\label{fig:bc_scatter_VO2_bodycomp_reversal}
\end{figure}
%----------------------------------------------------------------------------------------------

Absolute oxygen consumption (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$). However, this correlation was lost after correction of oxygen consumption for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.

Absolute oxygen consumption (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold. However, when it was corrected for total body weight, there was a strong correlation between corrected oxygen consumption (mls/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).

The loss of the physiological relationship between VO$_2$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig. \ref{fig:bc_scatter_VO2_TAT}.

%VO$_2$ corrected to height seems to correlate best with BC and everything else
\clearpage

@@

The positive correlation between absolute oxygen consumption and skeletal muscle area is easily explained by the physiology of aerobic exercise. During periods of increased physical activity, the greater oxygen demand is primarily due to increased metabolic activity within the skeletal muscle.

Current convention is to report oxygen consumption measured at cardiopulmonary exercise testing according the following formula: 
\[Corrected\ VO_2 (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ VO_2\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]

In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between oxygen consumption at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.

@@

Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{"Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation"}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass. In a detailed anaysis comparing oxygen consumption and body build, he concludes that \textit{"as the index wt./stature increases, O2/wt. must be expected to decrease purely as a result of the method used for representing the data." }

@@

Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998.\parencite{janz_longitudinal_1998} Aerobic capacity in the form of VO$_2$peak was evaluated annually in 126 children (mean age 10.3 years) for five years. Body composition changes were also tracked over this period. They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls. They reported on the significant difficulties in normalising VO$_2$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising VO$_2$. They found that VO$_2$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls. However, this underestimation was greater in girls than in boys.

Goran et al reported that total body fat did not affect maximal aerobic capacity.\parencite{goran_total_2000} They reported on VO$_2$max in obese women before and after weight loss. VO$_2$max corrected for total body weight was significantly lower in the obese state while VO$_2$max corrected for fat-free mass did not change significantly after weight loss. They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities. This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may increase the exercise load.

@@

In a study aimed at determining the optimal method of expressing VO$_2$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage.\parencite{maciejczyk_influence_2014} They reported that high body mass regardless of composition was correlated negatively with VO$_2$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat. However, when VO$_2$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. They, similar to Goran et al\parencite{goran_total_2000}, recommend that VO$_2$ be normalised to lean body mass rather than total body weight.

@@

\subsection{Clinical implications of spurious correlation}
Older et al in their pioneering study in 1993 reported that VO$_2$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993} While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a VO$_2$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between VO$_2$AT < 11mls/kg/min and mortality due to cardiovascular causes. \parencite{older_cardiopulmonary_1999} The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}

Swart and Carlisle reported that VO$_2$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes.\parencite{swart_case-controlled_2012} However, the proportion of females in the low VO$_2$AT group was significantly greater than that in the normal VO$_2$AT groups (24\% vs 51\%). The average VO$_2$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min. In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low VO$_2$AT group was 51\% while it was 28\% in the group with normal AT.\parencite{wilson_impaired_2010} There was no data presented on body mass index in this study.

@@

The design of such prehabilitation programs must not depend solely on body weight adjusted parameters of cardiopulmonary exercise testing when assessing the success of the interventions in these programs. Instead, improvement in the absolute values of VO$_2$AT and VO$_2$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.

>>>>

\clearpage
%----------------------------------------------------------------------------------------
$\dot{V}_{O_2}$
\section{Introduction}
Major abdominal surgery especially for pancreatic disease is associated with significant morbidity and mortality. Patient selection is as important as identifying surgical treatable pathology in ensuring optimal outcomes. \parencite{balthazar_acute_2002}

\subsection{Role of preoperative CPET}
The role of cardiopulmonary exercise testing in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery has become well established. A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery.\parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011} CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
Patients are sometimes denied surgery if their performance at cardiopulmonary exercise testing is felt to be poor based on currently available evidence.

@@

\subsection{Factors influencing aerobic fitness}
A low $\dot{V}_{O_2}$ has universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems to increased oxygen demand during exercise. This is often thought to be due to cardiorespiratory disease, overt or sub-clinical. Occasionally, other factors such as anaemia, peripheral vascular disease and rarely, mitochondrial diseases have been recognised as factors contributing to low $\dot{V}_{O_2}$ or abnormalities in other parameters measured at cardiopulmonary exercise testing but these causes are less common in patients undergoing major abdominal surgery.

The most common parameters used to quantify perioperative risk in surgical patients are oxygen consumption at the anaerobic threshold ($\dot{V}_{O_2}$AT) and at peak exercise capacity ($\dot{V}_{O_2}$Peak). Conventionally these have been reported as per weight ratios (ml/kg/min) to allow comparison between patients. However, numerous studies on cardiorespiratory exercise physiology have reported that normalising $\dot{V}_{O_2}$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects.\parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}

\subsection{Aims}
In chapter \ref{ch_cpet_outcomes}, we reported that low $\dot{V}_{O_2}$AT in patients undergoing pancreaticoduodenectomy was associated with increased incidence of postoperative pancreatic fistula and prolonged hospital stay. We also reported that patients with a $\dot{V}_{O_2}$AT less than 10 ml/kg/min were less likely to receive postoperative adjuvant chemotherapy as a result of postoperative complications, prolonged hospital stay and possibly due to lack of physiological reserve after surgery. However, we also observed that high BMI was associated with a low $\dot{V}_{O_2}$AT independent of all other clinicopathological characteristics. Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain the very low levels of $\dot{V}_{O_2}$AT.

The aim of the present study was to explore the association between body composition, total body weight and the physiological parameters measured at cardiopulmonary exercise testing.

@@

\subsection{Patients}
Patients scheduled to undergo pancreaticoduodenectomy for malignant or benign disease involving the head of the pancreas and periampullary region between August 2008 and October 2010 were included in this study. All data were recorded in a prospectively maintained database. Data was collected on demographics, preoperative clinicopathological characteristics including blood tests, body mass index, weight, height and the underlying surgical pathology. Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database. Cardiopulmonary exercise testing methodolgy is discussed in Section \ref{sec:cpx_method} (p\pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (p\pageref{sec:cpx_parameters}).

\subsection{Calculation of body composition}
Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published and well established methods.\parencite{bredella_comparison_2010,shen_total_2004}

%http://regionstraumapro.com/post/16349545265 - Excellent description of CT window width(W) and level/center(C)
The coronal and saggital reconstructions were used to accurately identify the L3 and L4 vertebrae. The CT window width (W) was set at 400 and the level/center (C) was set at 40. This allowed tissue between -160 HU and +240 HU to be represented in grayscale with adequate contrast between tissues of interest. The cross-sectional images at these levels where then exported as gray-scale bitmap images. The scale in millimeters was included with every image. A representative image is shown in Fig. 1.The GNU Image Manipulation Program (GIMP), an advanced, free, open-source, raster graphics editor was used for analysis of all images (www.gimp.org). The use of GIMP to analyse cross-sectional imaging for body composition has been described previously although by using a different technique to what has been employed by us. \parencite{anblagan_measurement_2013}

The first step involved converting the bitmap images into JPEG images using lossy compression set at 85\% to minimise sharp transitions between grey areas of very similar colour values. This allowed more consistent selection of contiguous areas of similar grey shades.

@@

\textbf{Selecting tissue compartments}:

The sequence of steps followed to calculate the area of each tissue compartment is depicted in Fig. \ref{fig:bc_ct_gimp} on p	\pageref{fig:bc_ct_gimp}. The total cross-sectional area of the abdomen at the level of L3/L4 was calculated by first selecting all the empty space outside the image followed by inverting this selection. This is depicted in Fig. \ref{fig:bc_ct_csa}. Subcutaneous fat in the image was selected using the Fuzzy Select tool (if necessary by choosing multiple times and removing any unnecessary areas) as depicted in Fig. \ref{fig:bc_ct_sat}. The same process was repeated for visceral adipose tissue and skeletal muscle as depicted in Fig. \ref{fig:bc_ct_vat} and Fig. \ref{fig:bc_ct_sm} respectively. Every selection was visually confirmed for anatomical accuracy by using the layer selection tool to inspect the area under selection as shown in the insets in each of the images.
%CT slices images x 4
\begin{figure}[htbp]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_csa}
		\caption{Total Cross-sectional Area}
		\label{fig:bc_ct_csa}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_sat}
		\caption{Subcutaneous Adipose Tissue$^*$}
		\label{fig:bc_ct_sat}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_vat}
		\caption{Visceral Adipose Tissue$^*$}
		\label{fig:bc_ct_vat}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_sm}
		\caption{Skeletal Muscle$^*$}
		\label{fig:bc_ct_sm}
	\end{subfigure}

@@

\subsection{Cardiopulmonary exercise testing}
All patients performed cardiopulmonary exercise testing on a cycle ergometer as described in Section \ref{sec:cpx_method}. Raw data of all breath-by-breath parameters averaged every 10 seconds was collected for analysis. The first three minutes of the recorded data were during the rest period when the patients were on the exercise bike but did not exercise. The average of each parameter measured between the first and second minute (6 readings) was treated as the rest value. Anaerobic threshold was identified using previously established methods and all corresponding parameters at this point were recorded.\parencite{beaver_new_1986,sue_metabolic_1988} Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.

@@

All analyses were performed using the SPSS statistical package for Microsoft Windows (version 22 ). Comparison between body composition and cardiopulmonary exercise testing parameters was done using Partial correlations controlling for the effect of gender (and/or age). All p-values reported are two-sided. The relationship between body composition and preoperative clinico-pathological characteristics (as categorical variables) was analysed using the Mann-Whitney U test for variables with two categories or the Kruskal-Wallis test for variables with more than two categories. Previously established cut-offs were used for categorising continuous variables where applicable. The level of significance was set at $p<0.05$.

@@

The correlations between cardiopulmonary exercise parameters and body composition were adjusted for gender. Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age. The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).

@@

Tidal volume ($\dot{V}_T$, litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise. There was a statistically significant but weak positive correlation between minute ventilation ($\dot{V}_E$, litres) and skeletal muscle at anaerobic threshold and peak exercise but not at rest. There was no correlation between either of these measures of pulmonary function and total adipose tissue area at any phase of exercise.

%----------------------------------------------------------------------------------------------
% Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal}
		\caption{$\dot{V}_{O_2}$AT vs. Skeletal Muscle}
		\label{fig:bc_scatter_VO2_skeletal}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT}
		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
		\label{fig:bc_scatter_VO2_TAT}
	\end{subfigure}
	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.}	
	\label{fig:bc_scatter_VO2_bodycomp_reversal}
\end{figure}
%----------------------------------------------------------------------------------------------

Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$). However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.

Absolute $\dot{V}_{O_2}$ (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold. However, when it was corrected for total body weight, there was a strong correlation between corrected $\dot{V}_{O_2}$ (mls/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).

The loss of the physiological relationship between $\dot{V}_{O_2}$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig. \ref{fig:bc_scatter_VO2_TAT}.

%$\dot{V}_{O_2}$ corrected to height seems to correlate best with BC and everything else
\clearpage

@@

The positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area is easily explained by the physiology of aerobic exercise. During periods of increased physical activity, the greater oxygen demand is primarily due to increased metabolic activity within the skeletal muscle.

Current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according the following formula: 
\[Corrected\ VO_2 (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ VO_2\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]

In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between $\dot{V}_{O_2}$ at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.

@@

Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{"Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation"}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass. In a detailed anaysis comparing $\dot{V}_{O_2}$ and body build, he concludes that \textit{"as the index wt./stature increases, O2/wt. must be expected to decrease purely as a result of the method used for representing the data." }

@@

Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998.\parencite{janz_longitudinal_1998} Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years. Body composition changes were also tracked over this period. They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls. They reported on the significant difficulties in normalising $\dot{V}_{O_2}$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising $\dot{V}_{O_2}$. They found that $\dot{V}_{O_2}$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls. However, this underestimation was greater in girls than in boys.

Goran et al reported that total body fat did not affect maximal aerobic capacity.\parencite{goran_total_2000} They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss. $\dot{V}_{O_2}$max corrected for total body weight was significantly lower in the obese state while $\dot{V}_{O_2}$max corrected for fat-free mass did not change significantly after weight loss. They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities. This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may increase the exercise load.

@@

In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage.\parencite{maciejczyk_influence_2014} They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat. However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. They, similar to Goran et al\parencite{goran_total_2000}, recommend that $\dot{V}_{O_2}$ be normalised to lean body mass rather than total body weight.

@@

\subsection{Clinical implications of spurious correlation}
Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993} While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes. \parencite{older_cardiopulmonary_1999} The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}

Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes.\parencite{swart_case-controlled_2012} However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%). The average $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min. In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT.\parencite{wilson_impaired_2010} There was no data presented on body mass index in this study.

@@

The design of such prehabilitation programs must not depend solely on body weight adjusted parameters of cardiopulmonary exercise testing when assessing the success of the interventions in these programs. Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.

%This needs to be expanded significantly.

%Mention studies that have used VE/VCO2. 
%I think there is a study in bariatrics where bmi and at were independent predictors. 
%Mention studies that used predicted VO2 or absolute $\dot{V}_{O_2}$ where calculated AT was not useful. 
%This should give me at least another 2 paragraphs

### 2015-06-06 Chapters/Introduction.tex
Chapter Body Composition - numerous small changes - fixed all VO2 to correct format - fixed some problems with style - fixed a few spelling/grammar issues - needs a final check - see comments at end of discussion - consider separate tables for each phase of exercise

\subsection{Cardiopulmonary Exercise Test Methodology}
CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.

@@

\section{Description of CPET Parameters}
\input{Tables/c1_cpet_parameters}

>>>>

\subsection{Cardiopulmonary Exercise Test Methodology}
\label{sec:cpx_method}
CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.

@@

\section{Description of CPET Parameters}
\label{sec:cpx_parameters}
\input{Tables/c1_cpet_parameters}
